---
title: "Summary statistics for bcftools stats results"
author: "CIG @ MGI"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
---

```{r config, include=FALSE}
# Repository location, separate from DATA_ROOT. Replace with the absolute path
# to your clone (the directory containing shared-R).
REPO_ROOT <- "/path/to/repo"
REPO_ROOT <- normalizePath(REPO_ROOT, mustWork = TRUE)
# Input locations. Set these before knitting.
DATA_ROOT <- "" # project data directory (per-analysis subdirectories)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

stats_dir <- file.path(DATA_ROOT, "Variant_Calling_Benchmark/Stats_Results")

knitr::opts_knit$set(root.dir = stats_dir)
set.seed(0)

require("ggplot2")
require("cowplot")
require("viridis")
require("reshape2")
require("dplyr")
require("tidyverse")
library(plotly)
library(Rtsne)
library(stringr)
library(ggbeeswarm)
library(ggforce)
library(ggbreak)
library(readr)

source(file.path(REPO_ROOT, "shared-R", "JFMV_2025_Aesthetic.R"))
```

```{r reader}
read_bcftools_stats <- function(file,
                                sections = c("SN","TSTV","SiS","AF","QUAL","IDD","ST","DP")) {
  lines <- readLines(file)
  stats <- list()
  for(sec in sections) {
    sec_lines <- grep(paste0("^", sec, "\t"), lines, value = TRUE)
    if(length(sec_lines) == 0) next
    df <- read.table(
      text             = sub(paste0("^", sec, "\t"), "", sec_lines),
      sep              = "\t",
      header           = FALSE,
      stringsAsFactors = FALSE,
      fill             = TRUE
    )
    stats[[sec]] <- df
  }
  stats
}
```

```{r load all files}
files <- list.files(stats_dir, pattern = "\\.stats$", full.names = TRUE)

sample_meta <- tibble(
  file     = files,
  basename = tools::file_path_sans_ext(basename(files))
) %>%
  mutate(
    method = str_extract(basename, "^[^_]+"),
    strain = str_extract(basename, "CC[0-9]+"),
    method = recode(method,
                    "bwa"                         = "bwa",
                    "Founders-surjected-chr-sort" = "graph-surjected")
  )

all_stats <- map(sample_meta$file, read_bcftools_stats)
names(all_stats) <- sample_meta$basename

bind_section <- function(section, colnames_vec) {
  map_dfr(names(all_stats), function(s) {
    df <- all_stats[[s]][[section]]
    if(is.null(df) || nrow(df) == 0) return(NULL)
    colnames(df) <- colnames_vec[seq_len(ncol(df))]
    df$sample <- s
    df
  }) %>%
    left_join(sample_meta %>% select(basename, method, strain),
              by = c("sample" = "basename"))
}

sn_all <- bind_section("SN", c("ID","key","value")) %>%
  mutate(
    key   = gsub("number of ", "", key),
    key   = gsub(":", "", key),
    value = suppressWarnings(as.numeric(value))
  ) %>%
  filter(!key %in% c("samples","no-ALTs"))

tstv_all <- bind_section(
  "TSTV",
  c("ID","ts","tv","tstv","ts_1st","tv_1st","tstv_1st")
) %>%
  mutate(across(c(ts, tv, tstv, ts_1st, tv_1st, tstv_1st),
                ~ suppressWarnings(as.numeric(.))))

af_all <- bind_section(
  "AF",
  c("ID","allele_frequency","number_SNPs","number_transitions",
    "number_transversions","number_indels",
    "repeat_consistent","repeat_inconsistent","not_applicable")
) %>%
  mutate(across(c(allele_frequency, number_SNPs, number_transitions,
                  number_transversions, number_indels),
                ~ suppressWarnings(as.numeric(.))))

idd_all <- bind_section("IDD", c("ID","length","count")) %>%
  mutate(length = suppressWarnings(as.numeric(length)),
         count  = suppressWarnings(as.numeric(count)))

st_all <- bind_section("ST", c("ID","type","count")) %>%
  mutate(count = suppressWarnings(as.numeric(count)))
```

# Overview

```{r overview}
sample_meta %>%
  count(method, name = "n_samples") %>%
  knitr::kable(caption = "Input files by method")
```

Loaded **`r nrow(sample_meta)`** `.stats` files spanning
**`r n_distinct(sample_meta$strain)`** strains and
**`r n_distinct(sample_meta$method)`** methods.

# Cross-sample comparisons {.tabset .tabset-fade}

## SN (summary numbers)

```{r sn compare, fig.width=12, fig.height=9}
ggplot(sn_all, aes(x = method, y = value, color = method)) +
  geom_quasirandom(size = 1.2, alpha = 0.8) +
  facet_wrap(~ key, scales = "free_y") +
  labs(x = NULL, y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

## Ts/Tv

```{r tstv compare, fig.width=8, fig.height=4}
tstv_long <- tstv_all %>%
  select(sample, method, strain, tstv, tstv_1st) %>%
  pivot_longer(c(tstv, tstv_1st), names_to = "type", values_to = "ratio")

ggplot(tstv_long, aes(x = method, y = ratio, color = method)) +
  geom_quasirandom(size = 1.5, alpha = 0.8) +
  facet_wrap(~ type) +
  labs(x = NULL, y = "Ts/Tv ratio") +
  theme(legend.position = "none")
```

## Allele frequency

```{r af compare, fig.width=12, fig.height=4}
af_long <- af_all %>%
  select(sample, method, strain, allele_frequency,
         number_SNPs, number_transitions, number_transversions, number_indels) %>%
  pivot_longer(c(number_SNPs, number_transitions,
                 number_transversions, number_indels),
               names_to = "metric", values_to = "count")

ggplot(af_long,
       aes(x = allele_frequency, y = count,
           color = method, group = sample)) +
  geom_line(alpha = 0.35) +
  facet_wrap(~ metric, nrow = 1, scales = "free_y") +
  labs(x = "Allele frequency", y = "Count") +
  theme(legend.position = "top")
```

## Indel length distribution

```{r idd compare, fig.width=10, fig.height=4}
ggplot(idd_all,
       aes(x = length, y = count, color = method, group = sample)) +
  geom_line(alpha = 0.35) +
  labs(x = "Length (deletions negative)", y = "Count") +
  theme(legend.position = "top")
```

## Substitution types

```{r st compare, fig.width=9, fig.height=4}
ggplot(st_all,
       aes(x = type, y = count / 1e6, color = method)) +
  geom_quasirandom(dodge.width = 0.8, size = 1.2, alpha = 0.8) +
  labs(x = "Substitution", y = "Count (x1e6)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
```

# PCA of stats {.tabset .tabset-fade}

```{r build feature matrix}
sn_wide <- sn_all %>%
  select(sample, key, value) %>%
  pivot_wider(names_from = key, values_from = value, names_prefix = "SN_")

tstv_wide <- tstv_all %>%
  select(sample, tstv, tstv_1st) %>%
  rename(TSTV_ratio = tstv, TSTV_ratio_1st = tstv_1st)

st_wide <- st_all %>%
  pivot_wider(id_cols = sample, names_from = type,
              values_from = count, names_prefix = "ST_",
              values_fn = sum)

idd_wide <- idd_all %>%
  mutate(bin = cut(length,
                   breaks = c(-Inf, -50, -10, -1, 1, 10, 50, Inf),
                   labels = c("del_gt50","del_10_50","del_1_10",
                              "ins_1_10","ins_10_50","ins_gt50","other"))) %>%
  group_by(sample, bin) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = bin, values_from = count,
              names_prefix = "IDD_", values_fill = 0)

feat <- sn_wide %>%
  left_join(tstv_wide, by = "sample") %>%
  left_join(st_wide,   by = "sample") %>%
  left_join(idd_wide,  by = "sample")

mat <- feat %>% column_to_rownames("sample") %>% as.matrix()
mat[is.na(mat)] <- 0
mat <- mat[, apply(mat, 2, function(x) sd(x, na.rm = TRUE) > 0)]

pca     <- prcomp(mat, scale. = TRUE)
var_exp <- summary(pca)$importance[2, ]

pca_df <- as.data.frame(pca$x) %>%
  rownames_to_column("sample") %>%
  left_join(sample_meta %>% select(basename, method, strain),
            by = c("sample" = "basename"))
```

## PC1 vs PC2

```{r pca12, fig.width=8, fig.height=6}
p <- ggplot(pca_df,
            aes(x = PC1, y = PC2, color = method,
                text = paste0(strain, " (", method, ")"))) +
  geom_point(size = 3, alpha = 0.85) +
  labs(x = sprintf("PC1 (%.1f%%)", var_exp[1] * 100),
       y = sprintf("PC2 (%.1f%%)", var_exp[2] * 100))
plotly::ggplotly(p, tooltip = "text")
```

## PC2 vs PC3

```{r pca23, fig.width=8, fig.height=6}
p <- ggplot(pca_df,
            aes(x = PC2, y = PC3, color = method,
                text = paste0(strain, " (", method, ")"))) +
  geom_point(size = 3, alpha = 0.85) +
  labs(x = sprintf("PC2 (%.1f%%)", var_exp[2] * 100),
       y = sprintf("PC3 (%.1f%%)", var_exp[3] * 100))
plotly::ggplotly(p, tooltip = "text")
```

## Scree

```{r scree, fig.width=6, fig.height=4}
data.frame(PC = seq_along(var_exp), var = var_exp) %>%
  head(15) %>%
  ggplot(aes(x = PC, y = var)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(y = "Variance explained")
```

## Top loadings

```{r loadings}
as.data.frame(pca$rotation[, 1:3]) %>%
  rownames_to_column("feature") %>%
  arrange(desc(abs(PC1))) %>%
  head(25) %>%
  knitr::kable(digits = 3, caption = "Top 25 features by |PC1 loading|")
```

# Per-sample reports {.tabset .tabset-dropdown}

```{r per-sample, results='asis', fig.width=10, fig.height=3}
for (s in sort(sample_meta$basename)) {
  cat("\n\n## ", s, "\n\n", sep = "")

  sn_s <- sn_all %>% filter(sample == s) %>% select(key, value)
  if (nrow(sn_s) > 0) {
    print(knitr::kable(sn_s, caption = "Summary numbers"))
    cat("\n\n")
  }

  af_s <- af_all %>% filter(sample == s)
  if (nrow(af_s) > 0) {
    p_af <- ggplot(af_s, aes(x = allele_frequency, y = number_SNPs)) +
      geom_col(fill = "steelblue") +
      labs(title = "AF: number of SNPs", x = "AF", y = "SNPs")
    print(p_af)
    cat("\n\n")
  }

  idd_s <- idd_all %>% filter(sample == s)
  if (nrow(idd_s) > 0) {
    p_idd <- ggplot(idd_s, aes(x = length, y = count)) +
      geom_col(fill = "steelblue") +
      labs(title = "Indel length distribution",
           x = "Length", y = "Count")
    print(p_idd)
    cat("\n\n")
  }

  st_s <- st_all %>% filter(sample == s)
  if (nrow(st_s) > 0) {
    p_st <- ggplot(st_s, aes(x = type, y = count)) +
      geom_col(fill = "steelblue") +
      labs(title = "Substitution types", x = NULL, y = "Count") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    print(p_st)
    cat("\n\n")
  }
}
```
