---
title: "Alignment Metrics Comparison: GRCm38 · mT2T · Founder Pangenome"
author: "CIG @ MGI"
date: "`r Sys.Date()`"
output: html_document
---

```{r config, include=FALSE}
# Repository location, separate from DATA_ROOT. Replace with the absolute path
# to your clone (the directory containing shared-R).
REPO_ROOT <- "/path/to/repo"
REPO_ROOT <- normalizePath(REPO_ROOT, mustWork = TRUE)
# Input locations. Set these before knitting.
DATA_ROOT <- "" # project data directory (per-analysis subdirectories)
```

```{r setup, include=FALSE}
set.seed(0)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = file.path(DATA_ROOT, "Alignment_Benchmarking/"))

library(ggplot2)
library(cowplot)
library(scales)
library(dplyr)
library(tidyr)
library(ggrepel)
library(ggbeeswarm)
library(data.table)
library(stringr)
library(patchwork)
source(file.path(REPO_ROOT, "shared-R", "JFMV_2025_Aesthetic.R"))

# NULL/NA-coalescing operator used by row builders
`%||%` <- function(a, b) {
    if (is.null(a) || length(a) == 0) return(b)
    if (is.atomic(a) && length(a) == 1 && is.na(a)) return(b)
    a
}
```

# Configuration

```{r analysis-settings}
METRICS_DIR <- "Metrics"
FIG_DIR     <- "figures"
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)

condition_order <- c(
    "GRCm38 + BWA",
    "GRCm38 + Giraffe (pre-surj)",
    "GRCm38 + Giraffe (surj)",
    "mT2T + BWA",
    "mT2T + Giraffe (pre-surj)",
    "mT2T + Giraffe (surj)",
    "Founders Pangenome (pre-surj)",
    "Founders Pangenome (surj)"
)

condition_palette <- c(
    "GRCm38 + BWA"                   = "#4a7c8a",
    "GRCm38 + Giraffe (pre-surj)"    = "#2e5966",
    "GRCm38 + Giraffe (surj)"        = "#8bbcc4",
    "mT2T + BWA"                     = "#d4883a",
    "mT2T + Giraffe (pre-surj)"      = "#c44e52",
    "mT2T + Giraffe (surj)"          = "#e8998d",
    "Founders Pangenome (pre-surj)"  = "#6a5acd",
    "Founders Pangenome (surj)"      = "#b0a8d8"
)

condition_short <- c(
    "GRCm38 + BWA"                   = "GRCm38\nBWA",
    "GRCm38 + Giraffe (pre-surj)"    = "GRCm38\nGir (pre)",
    "GRCm38 + Giraffe (surj)"        = "GRCm38\nGir (surj)",
    "mT2T + BWA"                     = "mT2T\nBWA",
    "mT2T + Giraffe (pre-surj)"      = "mT2T\nGir (pre)",
    "mT2T + Giraffe (surj)"          = "mT2T\nGir (surj)",
    "Founders Pangenome (pre-surj)"  = "Founders\n(pre)",
    "Founders Pangenome (surj)"      = "Founders\n(surj)"
)

# pre-surjection conditions (used by vg_long, vg_events, vg panels)
vg_conds <- c(
    "GRCm38 + Giraffe (pre-surj)",
    "mT2T + Giraffe (pre-surj)",
    "Founders Pangenome (pre-surj)"
)
```

# Parsers

```{r parsers}
# samtools stats: pull SN lines
parse_samtools_stats <- function(path) {
    lines <- readLines(path)
    sn <- grep("^SN\t", lines, value = TRUE)
    kv <- str_match(sn, "^SN\t([^:]+):\t([^\t#]+)")
    vals <- suppressWarnings(as.numeric(trimws(kv[, 3])))
    setNames(as.list(vals), trimws(kv[, 2]))
}

# samtools flagstat: primary, mapped, paired, singletons, etc.
parse_flagstats <- function(path) {
    lines <- readLines(path)
    out <- list()
    get_count <- function(pattern) {
        hit <- grep(pattern, lines, value = TRUE)
        if (length(hit) == 0) return(NA_real_)
        as.numeric(sub("^(\\d+).*", "\\1", hit[1]))
    }
    get_pct <- function(pattern) {
        hit <- grep(pattern, lines, value = TRUE)
        if (length(hit) == 0) return(NA_real_)
        m <- regmatches(hit[1], regexpr("\\(([0-9.]+)%", hit[1]))
        if (length(m) == 0) return(NA_real_)
        as.numeric(sub("\\(([0-9.]+)%.*", "\\1", m))
    }
    out$total           <- get_count("in total")
    out$primary         <- get_count("^[0-9]+ \\+ [0-9]+ primary$")
    out$primary_mapped  <- get_count("primary mapped")
    out$primary_mapped_pct <- get_pct("primary mapped")
    out$properly_paired <- get_count("properly paired")
    out$properly_paired_pct <- get_pct("properly paired")
    out$singletons      <- get_count("singletons")
    out$singletons_pct  <- get_pct("singletons")
    out$diff_chr_mapq5  <- get_count("with mate mapped to a different chr \\(mapQ>=5\\)")
    out
}

# custom-metrics.txt: primary/mapped/paired + MAPQ60/gapless/perfect
parse_custom_metrics <- function(path) {
    lines <- readLines(path)
    pick <- function(pattern) {
        hit <- grep(pattern, lines, value = TRUE)
        if (length(hit) == 0) return(NA_real_)
        as.numeric(sub(".*:\\s*", "", hit[1]))
    }
    list(
        total_primary       = pick("^Total primary reads"),
        mapped_reads        = pick("^Mapped reads"),
        properly_paired     = pick("^Properly paired reads"),
        mapq60              = pick("^MAPQ = 60 reads"),
        gapless             = pick("^Gapless reads"),
        perfect             = pick("^Perfect reads")
    )
}

# vg stats: pre-surjection graph metrics
parse_vg_stats <- function(path) {
    lines <- readLines(path)
    pick <- function(pattern) {
        hit <- grep(pattern, lines, value = TRUE)
        if (length(hit) == 0) return(NA_real_)
        as.numeric(sub(".*:\\s*([0-9]+).*", "\\1", hit[1]))
    }
    mq_line <- grep("^Mapping quality:", lines, value = TRUE)
    as_line <- grep("^Alignment score:", lines, value = TRUE)
    mq_mean <- if (length(mq_line)) as.numeric(sub(".*mean ([0-9.]+).*", "\\1", mq_line)) else NA_real_
    mq60    <- if (length(mq_line)) as.numeric(sub(".*max\\s+\\d+\\s*\\((\\d+).*", "\\1", mq_line)) else NA_real_
    as_mean <- if (length(as_line)) as.numeric(sub(".*mean ([0-9.]+).*", "\\1", as_line)) else NA_real_
    list(
        total_alignments       = pick("^Total alignments"),
        total_primary          = pick("^Total primary"),
        total_aligned          = pick("^Total aligned"),
        total_perfect          = pick("^Total perfect"),
        total_gapless          = pick("^Total gapless"),
        total_properly_paired  = pick("^Total properly paired"),
        mapq_mean              = mq_mean,
        mapq60                 = mq60,
        aln_score_mean         = as_mean,
        substitutions_bp       = pick("^Substitutions"),
        insertions_bp          = pick("^Insertions"),
        deletions_bp           = pick("^Deletions"),
        softclips_bp           = pick("^Softclips")
    )
}
```

# Row builders

```{r builders}
# For samtools-stats-based conditions (BWA and post-surj Giraffe/Founders)
row_samtools <- function(sample, condition, stats, flags, custom) {
    total <- stats[["raw total sequences"]]
    if (is.null(total) || is.na(total) || total == 0) total <- 1
    data.table(
        sample               = sample,
        condition            = condition,
        total_reads          = total,
        mapping_rate         = 100 * (stats[["reads mapped"]] %||% 0) / total,
        properly_paired_rate = 100 * (stats[["reads properly paired"]] %||% 0) / total,
        error_rate           = stats[["error rate"]] %||% NA_real_,
        mq0_rate             = 100 * (stats[["reads MQ0"]] %||% 0) / total,
        insert_size_avg      = stats[["insert size average"]] %||% NA_real_,
        mapq60_rate          = 100 * (custom$mapq60 %||% 0) / total,
        gapless_rate         = 100 * (custom$gapless %||% 0) / total,
        perfect_rate         = 100 * (custom$perfect %||% 0) / total,
        singletons_rate      = flags$singletons_pct %||% NA_real_,
        diff_chr_mapq5       = flags$diff_chr_mapq5 %||% NA_real_,
        aln_score_mean       = NA_real_,
        mapq_mean            = NA_real_,
        substitutions_bp     = NA_real_,
        insertions_bp        = NA_real_,
        deletions_bp         = NA_real_,
        softclips_bp         = NA_real_
    )
}

# For vg-stats-based conditions (pre-surj Giraffe/Founders)
row_vg <- function(sample, condition, vg) {
    total <- vg$total_primary
    if (is.null(total) || is.na(total) || total == 0) total <- 1
    data.table(
        sample               = sample,
        condition            = condition,
        total_reads          = total,
        mapping_rate         = 100 * (vg$total_aligned %||% 0) / total,
        properly_paired_rate = 100 * (vg$total_properly_paired %||% 0) / total,
        error_rate           = NA_real_,
        mq0_rate             = NA_real_,
        insert_size_avg      = NA_real_,
        mapq60_rate          = 100 * (vg$mapq60 %||% 0) / total,
        gapless_rate         = 100 * (vg$total_gapless %||% 0) / total,
        perfect_rate         = 100 * (vg$total_perfect %||% 0) / total,
        singletons_rate      = NA_real_,
        diff_chr_mapq5       = NA_real_,
        aln_score_mean       = vg$aln_score_mean,
        mapq_mean            = vg$mapq_mean,
        substitutions_bp     = vg$substitutions_bp %||% NA_real_,
        insertions_bp        = vg$insertions_bp %||% NA_real_,
        deletions_bp         = vg$deletions_bp %||% NA_real_,
        softclips_bp         = vg$softclips_bp %||% NA_real_
    )
}

```

# Load data

```{r load}
samples <- sort(unique(sub("\\.merged\\.stats\\.txt$", "",
                           list.files(file.path(METRICS_DIR, "GRCm38_BWA"),
                                      pattern = "\\.merged\\.stats\\.txt$"))))

safe <- function(expr) tryCatch(expr, error = function(e) NULL)

rows <- list()

for (s in samples) {
    # 1. GRCm38 + BWA
    p <- file.path(METRICS_DIR, "GRCm38_BWA")
    if (file.exists(file.path(p, paste0(s, ".merged.stats.txt")))) {
        st <- safe(parse_samtools_stats(file.path(p, paste0(s, ".merged.stats.txt"))))
        fl <- safe(parse_flagstats(file.path(p, paste0(s, ".merged.flagstats"))))
        cm <- safe(parse_custom_metrics(file.path(p, paste0(s, ".merged.custom-metrics.txt"))))
        if (!is.null(st)) rows[[length(rows) + 1]] <- row_samtools(s, "GRCm38 + BWA", st, fl, cm)
    }

    # 2. mT2T + BWA
    p <- file.path(METRICS_DIR, "mT2T_BWA")
    if (file.exists(file.path(p, paste0(s, ".merged.stats.txt")))) {
        st <- safe(parse_samtools_stats(file.path(p, paste0(s, ".merged.stats.txt"))))
        fl <- safe(parse_flagstats(file.path(p, paste0(s, ".merged.flagstats"))))
        cm <- safe(parse_custom_metrics(file.path(p, paste0(s, ".merged.custom-metrics.txt"))))
        if (!is.null(st)) rows[[length(rows) + 1]] <- row_samtools(s, "mT2T + BWA", st, fl, cm)
    }

    # 3. GRCm38 + Giraffe (surj)
    p <- file.path(METRICS_DIR, "GRCm38_Giraffe")
    stf <- file.path(p, paste0("md.renamed.sorted.", s, ".real.mm10.surjected.stats.txt"))
    flf <- file.path(p, paste0("md.renamed.sorted.", s, ".real.mm10.surjected.flagstats"))
    cmf <- file.path(p, paste0("md.renamed.sorted.", s, ".real.mm10.surjected.custom-metrics.txt"))
    if (file.exists(stf)) {
        st <- safe(parse_samtools_stats(stf))
        fl <- safe(parse_flagstats(flf))
        cm <- safe(parse_custom_metrics(cmf))
        if (!is.null(st)) rows[[length(rows) + 1]] <- row_samtools(s, "GRCm38 + Giraffe (surj)", st, fl, cm)
    }

    # 3b. GRCm38 + Giraffe (pre-surj)
    vgf <- file.path(p, paste0(s, ".real.mm10.stats.txt"))
    if (file.exists(vgf)) {
        vg <- safe(parse_vg_stats(vgf))
        if (!is.null(vg)) rows[[length(rows) + 1]] <- row_vg(s, "GRCm38 + Giraffe (pre-surj)", vg)
    }

    # 4. mT2T + Giraffe (surj); custom-metrics files carry a "corrected-" prefix
    p <- file.path(METRICS_DIR, "mT2T_Giraffe")
    stf <- file.path(p, paste0("corrected-md.renamed.sorted.", s, ".real.Yu-T2T.surjected.stats.txt"))
    flf <- file.path(p, paste0("md.renamed.sorted.", s, ".real.Yu-T2T.surjected.flagstats"))
    cmf <- file.path(p, paste0("corrected-md.renamed.sorted.", s, ".real.Yu-T2T.surjected.custom-metrics.txt"))
    if (file.exists(stf)) {
        st <- safe(parse_samtools_stats(stf))
        fl <- safe(parse_flagstats(flf))
        cm <- safe(parse_custom_metrics(cmf))
        if (!is.null(st)) rows[[length(rows) + 1]] <- row_samtools(s, "mT2T + Giraffe (surj)", st, fl, cm)
    }

    # 4b. mT2T + Giraffe (pre-surj)
    vgf <- file.path(p, paste0(s, ".real.Yu-T2T.stats.txt"))
    if (file.exists(vgf)) {
        vg <- safe(parse_vg_stats(vgf))
        if (!is.null(vg)) rows[[length(rows) + 1]] <- row_vg(s, "mT2T + Giraffe (pre-surj)", vg)
    }

    # 5. Founders (surj)
    p <- file.path(METRICS_DIR, "Founders")
    stf <- file.path(p, paste0("md.renamed.sorted.Founders.", s, ".surjected.stats.txt"))
    flf <- file.path(p, paste0("md.renamed.sorted.Founders.", s, ".surjected.flagstats"))
    cmf <- file.path(p, paste0("md.renamed.sorted.Founders.", s, ".surjected.custom-metrics.txt"))
    if (file.exists(stf)) {
        st <- safe(parse_samtools_stats(stf))
        fl <- safe(parse_flagstats(flf))
        cm <- safe(parse_custom_metrics(cmf))
        if (!is.null(st)) rows[[length(rows) + 1]] <- row_samtools(s, "Founders Pangenome (surj)", st, fl, cm)
    }

    # 5b. Founders (pre-surj)
    vgf <- file.path(p, paste0("Founders.", s, ".stats.txt"))
    if (file.exists(vgf)) {
        vg <- safe(parse_vg_stats(vgf))
        if (!is.null(vg)) rows[[length(rows) + 1]] <- row_vg(s, "Founders Pangenome (pre-surj)", vg)
    }
}

metrics_dt <- rbindlist(rows, use.names = TRUE, fill = TRUE)
metrics_dt[, condition := factor(condition, levels = condition_order, ordered = TRUE)]

# Sample counts per condition
metrics_dt[, .N, by = condition][order(condition)]
```

```{r vg_summary}
metrics_dt[condition %in% vg_conds, 
           .(mean = mean(aln_score_mean, na.rm = TRUE),
             sd = sd(aln_score_mean, na.rm = TRUE)),
           by = condition]
```

# Long format for faceted plotting

```{r long}
metric_cols <- c("mapping_rate", "properly_paired_rate", "perfect_rate",
                 "gapless_rate", "mapq60_rate", "error_rate",
                 "mq0_rate", "insert_size_avg", "singletons_rate")

metric_labels <- c(
    mapping_rate         = "Mapping rate (%)",
    properly_paired_rate = "Properly paired (%)",
    perfect_rate         = "Perfect reads (%)",
    gapless_rate         = "Gapless reads (%)",
    mapq60_rate          = "MAPQ = 60 reads (%)",
    error_rate           = "Error rate",
    mq0_rate             = "MQ0 reads (%)",
    insert_size_avg      = "Insert size (bp)",
    singletons_rate      = "Singletons (%)"
)

# Output combined per-sample metrics dataframe
fwrite(metrics_dt, "alignment_metrics_combined.tsv", sep = "\t")

metrics_long <- melt(
    metrics_dt,
    id.vars = c("sample", "condition", "total_reads"),
    measure.vars = metric_cols,
    variable.name = "metric",
    value.name = "value",
    na.rm = TRUE
)
metrics_long[, metric := factor(metric, levels = metric_cols)]
```

# Plotting function

```{r plot_fun}
plot_metric_box <- function(dt, metric_name, ylab, conds = NULL) {
    if (is.null(conds)) {
        conds <- intersect(condition_order, unique(as.character(dt$condition)))
    }
    sub <- dt[condition %in% conds & metric == metric_name]
    sub[, condition := factor(condition, levels = conds, ordered = TRUE)]

    medians <- sub[, .(median_value = median(value, na.rm = TRUE)), by = condition]

    ggplot(sub, aes(x = condition, y = value, fill = condition)) +
        geom_boxplot(
            outlier.shape = NA,
            width = 0.6,
            alpha = 0.4
        ) +
        geom_quasirandom(
            aes(color = condition),
            width = 0.18,
            alpha = 0.85,
            size = 1,
            stroke = 0
        ) +
        geom_point(
            data = medians,
            aes(x = condition, y = median_value),
            inherit.aes = FALSE,
            shape = 95,
            size = 8,
            color = "black"
        ) +
        scale_fill_manual(values = condition_palette, drop = FALSE) +
        scale_color_manual(values = condition_palette, drop = FALSE) +
        scale_x_discrete(labels = condition_short) +
        labs(x = NULL, y = ylab) +
        theme_cowplot() +
        JFMVTheme +
        theme(
            legend.position = "none",
            axis.text.x = element_text(size = 7, lineheight = 0.85)
        )
}
```

# Individual metric panels

## Mapping rate

```{r panel_mapping_rate, fig.width = 4, fig.height = 4}
p_mapping_rate <- plot_metric_box(metrics_long, "mapping_rate", "Mapping rate (%)")
p_mapping_rate
ggsave(file.path(FIG_DIR, "box_mapping_rate.svg"),
       p_mapping_rate, width = 4, height = 1.5, units = "in")
```

## Properly paired rate

```{r panel_properly_paired_rate, fig.width = 4, fig.height = 4}
p_properly_paired_rate <- plot_metric_box(metrics_long, "properly_paired_rate", "Properly paired (%)")
p_properly_paired_rate
ggsave(file.path(FIG_DIR, "box_properly_paired_rate.svg"),
       p_properly_paired_rate, width = 4, height = 1.5, units = "in")
```

## Perfect reads

```{r panel_perfect_rate, fig.width = 4, fig.height = 4}
p_perfect_rate <- plot_metric_box(metrics_long, "perfect_rate", "Perfect reads (%)")
p_perfect_rate
ggsave(file.path(FIG_DIR, "box_perfect_rate.svg"),
       p_perfect_rate, width = 4, height = 1.5, units = "in")
```

## Gapless reads

```{r panel_gapless_rate, fig.width = 4, fig.height = 4}
p_gapless_rate <- plot_metric_box(metrics_long, "gapless_rate", "Gapless reads (%)")
p_gapless_rate
ggsave(file.path(FIG_DIR, "box_gapless_rate.svg"),
       p_gapless_rate, width = 4, height = 1.5, units = "in")
```

## MAPQ = 60 rate

```{r panel_mapq60_rate, fig.width = 4, fig.height = 4}
p_mapq60_rate <- plot_metric_box(metrics_long, "mapq60_rate", "MAPQ = 60 reads (%)")
p_mapq60_rate
ggsave(file.path(FIG_DIR, "box_mapq60_rate.svg"),
       p_mapq60_rate, width = 4, height = 1.5, units = "in")
```

## Error rate

```{r panel_error_rate, fig.width = 4, fig.height = 4}
p_error_rate <- plot_metric_box(metrics_long, "error_rate", "Error rate")
p_error_rate
ggsave(file.path(FIG_DIR, "box_error_rate.svg"),
       p_error_rate, width = 4, height = 1.5, units = "in")
```

## MQ0 rate

```{r panel_mq0_rate, fig.width = 4, fig.height = 4}
p_mq0_rate <- plot_metric_box(metrics_long, "mq0_rate", "MQ0 reads (%)")
p_mq0_rate
ggsave(file.path(FIG_DIR, "box_mq0_rate.svg"),
       p_mq0_rate, width = 4, height = 1.5, units = "in")
```

## Insert size

```{r panel_insert_size, fig.width = 4, fig.height = 4}
p_insert_size <- plot_metric_box(metrics_long, "insert_size_avg", "Insert size (bp)")
p_insert_size
ggsave(file.path(FIG_DIR, "box_insert_size_avg.svg"),
       p_insert_size, width = 4, height = 1.5, units = "in")
```

## Singletons rate

```{r panel_singletons_rate, fig.width = 4, fig.height = 4}
p_singletons_rate <- plot_metric_box(metrics_long, "singletons_rate", "Singletons (%)")
p_singletons_rate
ggsave(file.path(FIG_DIR, "box_singletons_rate.svg"),
       p_singletons_rate, width = 4, height = 1.5, units = "in")
```

# VG-specific panels (pre-surjection only)

```{r vg_long}
vg_long <- melt(
    metrics_dt[condition %in% vg_conds],
    id.vars = c("sample", "condition"),
    measure.vars = c("aln_score_mean", "mapq_mean"),
    variable.name = "metric",
    value.name = "value",
    na.rm = TRUE
)
```

## Mean alignment score

```{r panel_aln_score_mean, fig.width = 4, fig.height = 4}
p_aln_score_mean <- plot_metric_box(vg_long, "aln_score_mean", "Mean alignment score", conds = vg_conds)
p_aln_score_mean
ggsave(file.path(FIG_DIR, "vg_aln_score_mean.svg"),
       p_aln_score_mean, width = 4, height = 1.5, units = "in")
```

## Mean mapping quality

```{r panel_mapq_mean, fig.width = 4, fig.height = 4}
p_mapq_mean <- plot_metric_box(vg_long, "mapq_mean", "Mean mapping quality", conds = vg_conds)
p_mapq_mean
ggsave(file.path(FIG_DIR, "vg_mapq_mean.svg"),
       p_mapq_mean, width = 4, height = 1.5, units = "in")
```

# Variant events (pre-surjection bp per read)

```{r vg_events, fig.width = 4, fig.height = 4}
events <- metrics_dt[
    condition %in% vg_conds,
    .(sample, condition,
      Substitutions = substitutions_bp / total_reads,
      Insertions    = insertions_bp / total_reads,
      Deletions     = deletions_bp / total_reads,
      Softclips     = softclips_bp / total_reads)
]
events_long <- melt(
    events,
    id.vars = c("sample", "condition"),
    variable.name = "event",
    value.name = "bp_per_read"
)
events_long[, event := factor(event, levels = c("Substitutions", "Insertions", "Deletions", "Softclips"))]

vg_events_plot <- ggplot(
    events_long,
    aes(x = event, y = bp_per_read, fill = condition)
) +
    geom_boxplot(
        outlier.shape = NA,
        position = position_dodge(0.8),
        width = 0.6,
        alpha = 0.5
    ) +
    geom_quasirandom(
        aes(color = condition),
        dodge.width = 0.8,
        width = 0.12,
        size = 0.6,
        alpha = 0.85,
        stroke = 0
    ) +
    scale_fill_manual(values = condition_palette) +
    scale_color_manual(values = condition_palette) +
    labs(x = NULL, y = "bp per read", fill = NULL, color = NULL) +
    theme_cowplot() +
    JFMVTheme +
    theme(legend.position = "bottom", legend.text = element_text(size = 7))

vg_events_plot
ggsave("figures/vg_variant_events.svg", vg_events_plot, width = 4, height = 1.5, units = "in")
```

# Pre vs post surjection scatter

```{r pp_scatter, fig.width = 4, fig.height = 4}
pp_pairs <- list(
    c("GRCm38 + Giraffe",        "GRCm38 + Giraffe (pre-surj)",     "GRCm38 + Giraffe (surj)"),
    c("mT2T + Giraffe",          "mT2T + Giraffe (pre-surj)",       "mT2T + Giraffe (surj)"),
    c("Founders Pangenome",      "Founders Pangenome (pre-surj)",   "Founders Pangenome (surj)")
)

pp_rows <- rbindlist(lapply(pp_pairs, function(trio) {
    label <- trio[1]; pre_c <- trio[2]; post_c <- trio[3]
    pre  <- metrics_dt[condition == pre_c, .(sample, pre_mapping  = mapping_rate,
                                             pre_perfect  = perfect_rate,
                                             pre_gapless  = gapless_rate,
                                             pre_paired   = properly_paired_rate)]
    post <- metrics_dt[condition == post_c, .(sample, post_mapping = mapping_rate,
                                               post_perfect = perfect_rate,
                                               post_gapless = gapless_rate,
                                               post_paired  = properly_paired_rate)]
    merged <- merge(pre, post, by = "sample")
    merged[, condition_group := label]
    merged
}))

scatter_metric <- function(pre_col, post_col, xlab, ylab) {
    ggplot(pp_rows, aes_string(x = pre_col, y = post_col, color = "condition_group")) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#999") +
        geom_point(alpha = 0.75, size = 1.5) +
        facet_wrap(~ condition_group, nrow = 1, scales = "free") +
        scale_color_manual(values = c(
            "GRCm38 + Giraffe"     = "#2e5966",
            "mT2T + Giraffe"       = "#c44e52",
            "Founders Pangenome"   = "#6a5acd"
        )) +
        labs(x = xlab, y = ylab) +
        theme_cowplot() +
        JFMVTheme +
        theme(legend.position = "none", strip.text = element_text(size = 8))
}

```

## Mapping rate pre vs post

```{r pp_scatter_mapping, fig.width = 12, fig.height = 4}
p_pp_mapping <- scatter_metric("pre_mapping", "post_mapping",
                               "Pre-surjection aligned (%)", "Post-surjection mapped (%)")
p_pp_mapping
ggsave(file.path(FIG_DIR, "pp_scatter_mapping.svg"),
       p_pp_mapping, width = 12, height = 4, units = "in")
```

## Perfect reads pre vs post

```{r pp_scatter_perfect, fig.width = 12, fig.height = 4}
p_pp_perfect <- scatter_metric("pre_perfect", "post_perfect",
                               "Pre-surjection perfect (%)", "Post-surjection perfect (%)")
p_pp_perfect
ggsave(file.path(FIG_DIR, "pp_scatter_perfect.svg"),
       p_pp_perfect, width = 12, height = 4, units = "in")
```

## Gapless reads pre vs post

```{r pp_scatter_gapless, fig.width = 12, fig.height = 4}
p_pp_gapless <- scatter_metric("pre_gapless", "post_gapless",
                               "Pre-surjection gapless (%)", "Post-surjection gapless (%)")
p_pp_gapless
ggsave(file.path(FIG_DIR, "pp_scatter_gapless.svg"),
       p_pp_gapless, width = 12, height = 4, units = "in")
```

## Properly paired pre vs post

```{r pp_scatter_paired, fig.width = 12, fig.height = 4}
p_pp_paired <- scatter_metric("pre_paired", "post_paired",
                              "Pre-surjection paired (%)", "Post-surjection paired (%)")
p_pp_paired
ggsave(file.path(FIG_DIR, "pp_scatter_paired.svg"),
       p_pp_paired, width = 12, height = 4, units = "in")
```

# Per-sample heatmaps

```{r heatmap_fun}
heatmap_metric <- function(metric_name, label, reverse = FALSE) {
    sub <- metrics_long[metric == metric_name]
    if (nrow(sub) == 0) return(invisible(NULL))

    pal <- if (reverse) rev(RColorBrewer::brewer.pal(11, "RdYlGn"))
           else         RColorBrewer::brewer.pal(11, "RdYlGn")

    ggplot(sub, aes(x = condition, y = sample, fill = value)) +
        geom_tile(color = "white", linewidth = 0.2) +
        scale_fill_gradientn(colours = pal, name = label) +
        scale_x_discrete(labels = condition_short) +
        labs(x = NULL, y = NULL, title = label) +
        theme_cowplot() +
        JFMVTheme +
        theme(
            axis.text.x = element_text(size = 7, lineheight = 0.85),
            axis.text.y = element_text(size = 6),
            legend.position = "right"
        )
}
```

## Mapping rate heatmap

```{r heatmap_mapping_rate, fig.width = 6, fig.height = 4}
hm_mapping_rate <- heatmap_metric("mapping_rate", "Mapping rate (%)")
hm_mapping_rate
ggsave(file.path(FIG_DIR, "heatmap_mapping_rate.svg"),
       hm_mapping_rate, width = 6, height = 4, units = "in")
```

## Properly paired heatmap

```{r heatmap_properly_paired, fig.width = 6, fig.height = 4}
hm_properly_paired <- heatmap_metric("properly_paired_rate", "Properly paired (%)")
hm_properly_paired
ggsave(file.path(FIG_DIR, "heatmap_properly_paired.svg"),
       hm_properly_paired, width = 6, height = 4, units = "in")
```

## Perfect reads heatmap

```{r heatmap_perfect, fig.width = 6, fig.height = 4}
hm_perfect <- heatmap_metric("perfect_rate", "Perfect reads (%)")
hm_perfect
ggsave(file.path(FIG_DIR, "heatmap_perfect.svg"),
       hm_perfect, width = 6, height = 4, units = "in")
```

## Gapless reads heatmap

```{r heatmap_gapless, fig.width = 6, fig.height = 4}
hm_gapless <- heatmap_metric("gapless_rate", "Gapless reads (%)")
hm_gapless
ggsave(file.path(FIG_DIR, "heatmap_gapless.svg"),
       hm_gapless, width = 6, height = 4, units = "in")
```

## MAPQ = 60 heatmap

```{r heatmap_mapq60, fig.width = 6, fig.height = 4}
hm_mapq60 <- heatmap_metric("mapq60_rate", "MAPQ = 60 (%)")
hm_mapq60
ggsave(file.path(FIG_DIR, "heatmap_mapq60.svg"),
       hm_mapq60, width = 6, height = 4, units = "in")
```

## Error rate heatmap

```{r heatmap_error, fig.width = 6, fig.height = 4}
hm_error <- heatmap_metric("error_rate", "Error rate", reverse = TRUE)
hm_error
ggsave(file.path(FIG_DIR, "heatmap_error.svg"),
       hm_error, width = 6, height = 4, units = "in")
```

# Summary table

```{r summary}
summary_dt <- metrics_long[,
    .(N = .N,
      mean = round(mean(value, na.rm = TRUE), 3),
      sd   = round(sd(value,   na.rm = TRUE), 3),
      median = round(median(value, na.rm = TRUE), 3)),
    by = .(condition, metric)
][order(metric, condition)]

summary_dt
fwrite(summary_dt, "alignment_metrics_summary.tsv", sep = "\t")
```

# Session info

```{r session}
sessionInfo()
```
