---
title: "Genomic Features at Personalization Mismatch Sites"
author: "CIG @ MGI"
date: "`r Sys.Date()`"
output: html_document
---

```{r config, include=FALSE}
# Repository location, separate from DATA_ROOT. Replace with the absolute path
# to your clone (the directory containing shared-R).
REPO_ROOT <- "/path/to/repo"
REPO_ROOT <- normalizePath(REPO_ROOT, mustWork = TRUE)
# Input locations. Set these before knitting.
DATA_ROOT <- "" # project data directory (per-analysis subdirectories)
```

```{r setup, include=FALSE}
set.seed(0)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = file.path(DATA_ROOT, "Low_Concordance_Regions/"))

library(ggplot2)
library(cowplot)
library(dplyr)
library(tidyr)
library(scales)
library(ggrepel)
source(file.path(REPO_ROOT, "shared-R", "JFMV_2025_Aesthetic.R"))
```

---

## 1  Load data

The rmsk BED files (mismatch_rmsk.bed, match_rmsk.bed) have one row per
(site × overlapping repeat) from bedtools intersect -loj. Columns:
1–7 from the site BED: chrom, start, end, var_type, n_founders_nonref, n_mismatch, n_match
8–13 from mm10_rmsk.bed: rmsk_chrom, rmsk_start, rmsk_end, repeat_name, repeat_class, repeat_family
("." in cols 8–13 when no overlap)

```{r load_data}
site_cols <- c(
  "chrom","start","end","var_type","n_founders_nonref","n_mismatch","n_match"
)
rmsk_cols <- c(
  "rmsk_chrom","rmsk_start","rmsk_end","repeat_name","repeat_class","repeat_family"
)
all_cols <- c(site_cols, rmsk_cols)

mm_rmsk <- read.delim("mismatch_rmsk.bed", header = FALSE,
                       col.names = all_cols, stringsAsFactors = FALSE)
mc_rmsk <- read.delim("match_rmsk.bed",    header = FALSE,
                       col.names = all_cols, stringsAsFactors = FALSE)

# Flag repeat overlap and group status
mm_rmsk <- mm_rmsk %>% mutate(in_repeat = repeat_class != ".", group = "Mismatch")
mc_rmsk <- mc_rmsk %>% mutate(in_repeat = repeat_class != ".", group = "Match")

# Site-level views (one row per site; for sites with multiple repeat hits,
# keep first hit: in_repeat status and class of the first overlapping repeat)
mm_sites <- mm_rmsk %>%
  distinct(chrom, start, end, .keep_all = TRUE)
mc_sites <- mc_rmsk %>%
  distinct(chrom, start, end, .keep_all = TRUE)

cat("Mismatch sites:", nrow(mm_sites), "\n")
cat("Match sites:   ", nrow(mc_sites), "\n")
```

---

## 2  Variant type distribution

```{r vartype_counts, fig.width=5, fig.height=4}
vt_mm <- mm_sites %>% count(var_type) %>% mutate(group = "Mismatch")
vt_mc <- mc_sites %>% count(var_type) %>% mutate(group = "Match")
vt    <- bind_rows(vt_mm, vt_mc) %>%
  group_by(group) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

vt_order <- vt %>% filter(group == "Mismatch") %>% arrange(desc(pct)) %>% pull(var_type)
vt$var_type <- factor(vt$var_type, levels = vt_order)

p_vt <- ggplot(vt, aes(x = var_type, y = pct, fill = group)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Mismatch" = "#d95f02", "Match" = "#1b9e77")) +
  labs(x = "Variant type", y = "% of sites", fill = NULL,
       title = "Variant type: Mismatch vs. Match sites") +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
p_vt
ggsave("../Vartype_mismatch_vs_match.svg", p_vt, width = 5, height = 4)
```

```{r vartype_foldenrich, fig.width=4.5, fig.height=4}
vt_wide <- vt %>%
  select(var_type, group, pct) %>%
  pivot_wider(names_from = group, values_from = pct, values_fill = 0) %>%
  mutate(log2_fe = log2(Mismatch / Match))

p_fe <- ggplot(vt_wide, aes(x = reorder(var_type, log2_fe), y = log2_fe,
                              fill = log2_fe > 0)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = "#d95f02", "FALSE" = "#1b9e77"), guide = "none") +
  labs(x = NULL, y = "log2(Mismatch% / Match%)",
       title = "Variant type fold-enrichment in mismatch sites") +
  coord_flip() +
  theme_cowplot()
p_fe
ggsave("../Vartype_foldenrich_mismatch.svg", p_fe, width = 4.5, height = 4)
```

```{r vartype_fisher}
fisher_vt <- lapply(unique(vt$var_type), function(vt_name) {
  n_mm_in  <- sum(mm_sites$var_type == vt_name)
  n_mm_out <- nrow(mm_sites) - n_mm_in
  n_mc_in  <- sum(mc_sites$var_type == vt_name)
  n_mc_out <- nrow(mc_sites) - n_mc_in
  ft <- fisher.test(matrix(c(n_mm_in, n_mm_out, n_mc_in, n_mc_out), nrow = 2))
  data.frame(var_type = vt_name, n_mismatch = n_mm_in, n_match = n_mc_in,
             odds_ratio = ft$estimate, p_value = ft$p.value)
}) %>%
  bind_rows() %>%
  mutate(padj = p.adjust(p_value, method = "BH")) %>%
  arrange(p_value)

print(fisher_vt)
```

---

## 3  Founder allele ambiguity (n_founders_nonref)

```{r founder_ambiguity, fig.width=5, fig.height=4}
amb <- bind_rows(
  mm_sites %>% count(n_founders_nonref) %>% mutate(group = "Mismatch"),
  mc_sites %>% count(n_founders_nonref) %>% mutate(group = "Match")
) %>%
  group_by(group) %>% mutate(pct = n / sum(n) * 100) %>% ungroup()

p_amb <- ggplot(amb, aes(x = factor(n_founders_nonref), y = pct, fill = group)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Mismatch" = "#d95f02", "Match" = "#1b9e77")) +
  labs(x = "# founders with non-ref allele (0–8)", y = "% of sites", fill = NULL,
       title = "Founder allele ambiguity: Mismatch vs. Match sites") +
  theme_cowplot()
p_amb
ggsave("../Founder_ambiguity_mismatch_vs_match.svg", p_amb, width = 5, height = 4)
```

```{r founder_ambiguity_stats}
wt <- wilcox.test(mm_sites$n_founders_nonref, mc_sites$n_founders_nonref)
cat(sprintf("Median n_founders_nonref. Mismatch: %.1f | Match: %.1f\n",
            median(mm_sites$n_founders_nonref), median(mc_sites$n_founders_nonref)))
cat(sprintf("Wilcoxon rank-sum: W = %.0f, p = %.2e\n", wt$statistic, wt$p.value))
```

---

## 4  RepeatMasker enrichment

```{r repeat_overall, fig.width=4, fig.height=3.5}
pct_mm_rep <- mean(mm_sites$in_repeat) * 100
pct_mc_rep <- mean(mc_sites$in_repeat) * 100

cat(sprintf("Mismatch sites in repeat: %.2f%%\n", pct_mm_rep))
cat(sprintf("Match    sites in repeat: %.2f%%\n", pct_mc_rep))

ft_rep <- fisher.test(matrix(
  c(sum(mm_sites$in_repeat),  nrow(mm_sites) - sum(mm_sites$in_repeat),
    sum(mc_sites$in_repeat),  nrow(mc_sites) - sum(mc_sites$in_repeat)),
  nrow = 2
))
cat(sprintf("Fisher OR = %.3f, p = %.2e\n", ft_rep$estimate, ft_rep$p.value))

p_ovr <- ggplot(
  data.frame(group = c("Mismatch","Match"), pct = c(pct_mm_rep, pct_mc_rep)),
  aes(x = group, y = pct, fill = group)
) +
  geom_col(width = 0.5) +
  scale_fill_manual(values = c("Mismatch" = "#d95f02", "Match" = "#1b9e77"), guide = "none") +
  labs(x = NULL, y = "% sites overlapping repeat", title = "Overall repeat overlap") +
  theme_cowplot() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
p_ovr
ggsave("../Repeat_overlap_overall.svg", p_ovr, width = 4, height = 3.5)
```

```{r repeat_class_enrichment, fig.width=6, fig.height=5}
# Count how many sites overlap each repeat class
# (use full rmsk table, one row per site×repeat hit; exclude "." rows)
rc_mm <- mm_rmsk %>% filter(repeat_class != ".") %>%
  distinct(chrom, start, end, repeat_class) %>%
  count(repeat_class) %>%
  mutate(pct = n / nrow(mm_sites) * 100, group = "Mismatch")

rc_mc <- mc_rmsk %>% filter(repeat_class != ".") %>%
  distinct(chrom, start, end, repeat_class) %>%
  count(repeat_class) %>%
  mutate(pct = n / nrow(mc_sites) * 100, group = "Match")

rc_wide <- bind_rows(rc_mm, rc_mc) %>%
  select(repeat_class, group, pct) %>%
  pivot_wider(names_from = group, values_from = pct, values_fill = 0) %>%
  mutate(
    log2_fe    = log2((Mismatch + 0.01) / (Match + 0.01)),
    total_pct  = Mismatch + Match
  ) %>%
  filter(total_pct > 0.05) %>%
  arrange(desc(log2_fe))

p_rc <- ggplot(rc_wide, aes(x = reorder(repeat_class, log2_fe), y = log2_fe,
                              fill = log2_fe > 0)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = "#d95f02", "FALSE" = "#1b9e77"), guide = "none") +
  labs(x = "Repeat class", y = "log2(Mismatch% / Match%)",
       title = "Repeat class enrichment in mismatch sites") +
  coord_flip() +
  theme_cowplot()
p_rc
ggsave("../Repeat_class_foldenrich.svg", p_rc, width = 6, height = 5)
```

```{r repeat_class_fisher}
all_classes <- rc_wide$repeat_class
n_mm <- nrow(mm_sites); n_mc <- nrow(mc_sites)

fisher_rc <- lapply(all_classes, function(cls) {
  n_in_mm  <- mm_rmsk %>% filter(repeat_class == cls) %>%
    distinct(chrom, start, end) %>% nrow()
  n_in_mc  <- mc_rmsk %>% filter(repeat_class == cls) %>%
    distinct(chrom, start, end) %>% nrow()
  ft <- fisher.test(matrix(
    c(n_in_mm, n_mm - n_in_mm, n_in_mc, n_mc - n_in_mc), nrow = 2
  ))
  data.frame(repeat_class = cls, n_mismatch = n_in_mm, n_match = n_in_mc,
             odds_ratio = ft$estimate, p_value = ft$p.value)
}) %>%
  bind_rows() %>%
  mutate(padj = p.adjust(p_value, method = "BH")) %>%
  arrange(p_value)

print(fisher_rc)
```

---

## 5  Combined: repeat × variant type

```{r repeat_by_vartype, fig.width=6, fig.height=4.5}
both_sites <- bind_rows(
  mm_sites %>% mutate(is_mismatch = TRUE),
  mc_sites %>% mutate(is_mismatch = FALSE)
)

rate_df <- both_sites %>%
  group_by(var_type, in_repeat) %>%
  summarise(
    n_total      = n(),
    n_mismatch   = sum(is_mismatch),
    mismatch_rate = n_mismatch / n_total,
    .groups = "drop"
  ) %>%
  filter(n_total > 100) %>%
  mutate(repeat_status = ifelse(in_repeat, "In repeat", "Not in repeat"))

p_combined <- ggplot(rate_df, aes(x = var_type, y = mismatch_rate * 100,
                                    fill = repeat_status)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("In repeat" = "#7570b3", "Not in repeat" = "#a6cee3")) +
  labs(x = "Variant type", y = "Mismatch rate (%)", fill = NULL,
       title = "Mismatch rate by variant type × repeat status") +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
p_combined
ggsave("../Mismatch_rate_vartype_x_repeat.svg", p_combined, width = 6, height = 4.5)
```

---

## 6  Genome-wide mismatch density (1Mb windows)

```{r genome_density, fig.width=12, fig.height=5}
chrom_levels <- paste0("chr", 1:19)

density_df <- both_sites %>%
  filter(chrom %in% chrom_levels) %>%
  mutate(
    chrom  = factor(chrom, levels = chrom_levels),
    window = floor(start / 1e6) * 1e6
  ) %>%
  group_by(chrom, window) %>%
  summarise(
    n_mismatch    = sum(is_mismatch),
    n_match       = sum(!is_mismatch),
    total_sites   = n(),
    mismatch_rate = n_mismatch / total_sites,
    .groups = "drop"
  ) %>%
  filter(total_sites >= 50)

p_dens <- ggplot(density_df, aes(x = window / 1e6, y = mismatch_rate * 100)) +
  geom_line(linewidth = 0.3, colour = "#d95f02", alpha = 0.8) +
  geom_smooth(method = "loess", span = 0.15, se = FALSE,
              colour = "black", linewidth = 0.6) +
  facet_wrap(~ chrom, nrow = 2, scales = "free_x") +
  labs(x = "Position (Mb)", y = "Mismatch rate (%)",
       title = "Genome-wide mismatch density (1Mb windows)") +
  theme_cowplot(font_size = 9) +
  theme(strip.text = element_text(size = 7))
p_dens
ggsave("../Genome_wide_mismatch_density.svg", p_dens, width = 12, height = 5)
```

---

## 7  Summary table

```{r summary_table}
data.frame(
  Feature = c(
    "Mismatch sites (autosomes, pooled)",
    "Match-only sites (autosomes, pooled)",
    "Overall mismatch rate (concordance pipeline)",
    "Mismatch sites in any repeat (%)",
    "Match sites in any repeat (%)",
    "Repeat enrichment OR (Fisher)",
    "Repeat enrichment p",
    "Median n_founders_nonref: mismatch",
    "Median n_founders_nonref: match",
    "Wilcoxon p (founder ambiguity)"
  ),
  Value = c(
    format(nrow(mm_sites), big.mark = ","),
    format(nrow(mc_sites), big.mark = ","),
    "0.81%",
    sprintf("%.2f%%", pct_mm_rep),
    sprintf("%.2f%%", pct_mc_rep),
    sprintf("%.3f", ft_rep$estimate),
    sprintf("%.2e", ft_rep$p.value),
    median(mm_sites$n_founders_nonref),
    median(mc_sites$n_founders_nonref),
    sprintf("%.2e", wt$p.value)
  )
) %>% knitr::kable(align = c("l","r"))
```
